Liveness-Based Garbage Collection

### 2014-11-10 paper/sections/liveness.tex
pushing liveness
Updated demand transformation
Updated Liveness Rules

\section{Liveness}\label{sec:liveness}

In classical liveness analysis a  variable is either `live' (its value
may be  used in future  computation) or `dead' (definitely  not used).
Semantically, a variable is dead at a given program point if arbitrary
changes to its value have no effect on the computation.  Later we will
use $\bot$ to represent a value  which `explodes' when it is used in a
computation; dead variables can  safely have their value replaced with
$\bot$.  For heap-allocated data we need a richer model of liveness in
that both variables  and fields of $\CONS$ cells may  be dead or live.
Using  $\acar$,  $\acdr$  to  represent access  using  $\CAR$,  $\CDR$
respectively, liveness of the structure reachable from a variable is a
set  of  {\em  access  paths}  which  we  represent  as  a  subset  of
$\{\acar,\acdr\}^\ast$, and  use conventional grammar  notation.  Thus
the liveness of $x$ being $\{\acdr\acar, \acdr\acdr\acar\}$ means that
future computation can  only refer to the second  and third members of
$x$   considered  as   a   list.   Semantically,   access  paths   are
prefix-closed, as  accessing a field requires accessing  all the paths
from  the variable  to  the field,  and  hence the  above liveness  is
properly   written   $\{\epsilon,   \acdr,   \acdr\acar,   \acdr\acdr,
\acdr\acdr\acar\}$.  The classical notions  of a scalar variable being
live or dead correspond to $\{\epsilon\}$ and $\{\}$.

@@

%==============================================================
\begin{figure}[t]  
% \footnotesize
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=& \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in \sigma\} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=& \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in \sigma\} \\
\mathit{ref\/}((\ID~x),\sigma,\Lfonly)
          &=& \{x.\sigma\} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=& \{x.\epsilon,y.\epsilon\} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=& \{x.\epsilon\} \\
\mathit{ref\/}((f~y_1~\cdots~y_n),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& x.\sigma \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \mathcal{L}(e_1,\sigma,\Lfonly) \cup
        \mathcal{L}(e_2,\sigma,\Lfonly) \cup
        \{x.\epsilon\} \\
\mathcal{L}((\LET~x \leftarrow  s~\IN~e),\sigma,\Lfonly) &=&
        \Lv
           \setminus \{x.*\}
           \cup \mathit{ref\/}(s,\Lv(x),\Lfonly)
\mbox{ where } \Lv = \mathcal{L}(e,\sigma,\Lfonly)
\end{eqnarray*}

\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}

\normalsize
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure}

>>>>

\section{Liveness}\label{sec:liveness}
%\comment{
To describe the issues that arise while formulating liveness of a lazy
language, consider the example shown in Figure ???

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{verbatim}
pi_0: let pi_1: x = (cons 1 x)
      in let  pi_2: y = (cdr x)
         in  let  pi_3: z = (car y)
             in pi_4: if (z > 0) then 
                       pi_5: let pi_6: w = (take 5 x)
                             in let pi_7: u = (cdr  w)
                                in l1:return u
                      else l2:return y

\end{verbatim}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The program is annotated with symbols standing for program points. For
reasons that will  become clear later, we use  different symbols for
the program points corresponding to return points (l instead of $\pi$)
Similar  to classical  liveness  analysis for  imperative programs,  a
demand  of  $\sigma$ on  the  expression  propagates  to a  demand  of
$\sigma$  on the return  values $u$  and $y$.  This gives  liveness of
$\lbrace u.\sigma \rbrace$ and $\{y.\sigma\}$ for the program points l1 and l2. We now
show how the liveness at l1 is propagated backwards; l2 can be treated
similarly.

Since u is (cdr w), a liveness of $\sigma$ for u translates into a
demand of $\sigma$ on $(cdr~w)$. This is turn translates into a liveness
of $\{\epsilon,~1.\sigma\}$ of $w$. In the case of eager evaluation, we
would have the livenesses of $\pi_7$ and l1 as $\{\epsilon,~1.\sigma\}$ and
${u.\sigma}$. The liveness of $u$ would have been killed at $\pi_7$. The
liveness of lazy evaluation differs in the following two respects:
\begin{enumerate}
\item The definition  $u = (cdr~w)$, does not  result in an evaluation
  of $(cdr~w)$; rather a  closure is created.  The evaluation actually
  takes place  while evaluating $(return~u)$. Thus it  is necessary to
  record the liveness of $\{\epsilon,  1.\sigma\}$ for $w$ also at the
  program point l1.

  Regard the  expression as a  tree and consider  paths from
  the  root  to   program  points  corresponding  to  return
  expressions.   The  above discussion  shows  that there  a
  single liveness  environment for each  such path. Further,
  if a program  point happens to fall on  more than one such
  path, then  the liveness environment of  the program point
  is the variable-wise union of the liveness environments of
  the individual paths.
\item The second difference is that since our lets are lazy,
  the let  definition can be  recursive as at  program point
  $\pi_1$. In such a case, the liveness has to be formulated
  as a fix-point computation.
\end{enumerate}

We therefore  formulate the per-program-point  liveness that
we require for garbage collection in terms of two ....

A  map $\cal{M}$  that gives  the common  liveness  for each
return path.

A  map that  gives  the  return paths  that  pass through  a
program point.

%}
In classical liveness analysis a  variable is either `live' (its value
may be  used in future  computation) or `dead' (definitely  not used).
Semantically, a variable is dead at a given program point if arbitrary
changes to its value have no effect on the computation.  Later we will
use $\bot$ to represent a value  which `explodes' when it is used in a
computation; dead variables can  safely have their value replaced with
$\bot$.  For heap-allocated data we need a richer model of liveness in
that both variables  and fields of $\CONS$ cells may  be dead or live.
Using  $\acar$,  $\acdr$  to  represent access  using  $\CAR$,  $\CDR$
respectively, liveness of the structure reachable from a variable is a
set  of  {\em  access  paths}  which  we  represent  as  a  subset  of
$\{\acar,\acdr\}^\ast$, and  use conventional grammar  notation.  Thus
the liveness of $x$ being $\{\acdr\acar, \acdr\acdr\acar\}$ means that
future computation can  only refer to the second  and third members of
$x$   considered  as   a   list.   Semantically,   access  paths   are
prefix-closed, as  accessing a field requires accessing  all the paths
from  the variable  to  the field,  and  hence the  above liveness  is
properly   written   $\{\epsilon,   \acdr,   \acdr\acar,   \acdr\acdr,
\acdr\acdr\acar\}$.  The classical notions  of a scalar variable being
live or dead correspond to $\{\epsilon\}$ and $\{\}$.

@@

%==============================================================
\begin{figure}[t]  
% \footnotesize
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& x.\sigma \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \cup
        \mathcal{L}(e_2,\sigma,\Lfonly) \cup
        \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \mathcal{L}(e_1,\sigma,\Lfonly) \cup
        \mathcal{L}(e_2,\sigma,\Lfonly)  & \mbox{otherwise}
                 \end{array} \\
        \mathcal{L}(\LET~\myvec{x} \leftarrow \myvec{s}~\IN~e),\sigma,\Lfonly) &=&
        \Lv
           \setminus \{\myvec{x}.*\}
           \cup LE(L)
\mbox{ where } \Lv = \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\myvec{x}}
\end{eqnarray*}

\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
\begin{minipage}{0.85\textwidth}
 \infrule[live-let-define]
         {f~\stackrel{\text{\tiny def}}{=}~(\DEFINE~(f~ x)~(\SIF~*~x~s[(f~ x)/x],~\mbox{and}~\Ldonly_{x \leftarrow s}~\stackrel{\text{\tiny def}}{=}~\Lfonly_f}       
         { x \leftarrow s \len \Ldonly_{x \leftarrow s}}
\end{minipage}
\normalsize
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure}
